NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2018-03-02 Manuscript/manuscriptPS.tex
Updates to manuscript

Based on the subjective ranking shown in Fig. \ref{comparisonTablePS},
the CHARMM36 is the best performing model for both lipids.
However, none of the tested models give a satisfactory 
agreement with experiments for the headgroup order paramaters of 
PS lipids. The total deviation from the experiments ($\Sigma$ in Fig. \ref{comparisonTablePS})
for the best performing CHARMM36 model is larger for PS lipids (8)
than for PC (3) \cite{botan15}. Therefore the 
interpretation of structural details of PS headgroup or
differences between PC and PS lipids is very challenging 
from the current MD simulation models.
Figures \ref{dihedralsGLY} and \ref{dihedralsHG} show dihedral
angle distributions calculated from different models. 
The glycerol backbone structures from CHARMM36 and
Slipids simulations visualized in Fig. \ref{glycerol_buslaev} reveal the
missing structures in Slipids model, which probably lead to the incorrect
order parameters. For the headgroup $\alpha$ and $\beta$ carbon order parameters of PS lipids,
the tested models perform less well than for PC headgroup in previous
study \cite{botan15}.
\todo{Discussion will be finished when we have all the data in
  Figs. \ref{HGorderParametersPS}, \ref{dihedralsGLY} and \ref{dihedralsHG}}

@@

The headgroup order parameters of PC lipids can be used to measure ion binding
affinity because they change proportionally
to the amount of bound charge in bilayer \cite{seelig87,catte16}.
This molecular electrometer concept can be applied also to
lipid bilayers with mixtures of PC and negatively charged
lipids \cite{borle85,macdonald87,roux90}.
This is demonstrated in Fig. %s \ref{OrderParametersWithCaCl},
%\ref{OrderParametersWithCaClBELOW1M} and
\ref{OrderParameterCHANGESWithCaClBELOW1M}
using the previuosly reported experimental data
for mixtures with varying proportions of
negatively charged PS or PG lipids.
%
%the changes of PC headgroup order parameters %for PC headgroup $\alpha$ and $\beta$ carbons
%as a function of CaCl$_2$ concentration are shown
%
As expected, the decrease of order parameters with CaCl$_2$
is more pronounced for systems with more
negatively charged lipids.
% (see Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}).

@@

%\begin{figure}[]
%  \centering
%  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaCl.eps}
%  \caption{\label{OrderParametersWithCaCl}
%    PC headgroup order parameters as a function of CaCl concentration from experiments containing charged lipids.
%    Pure DPPC data from \cite{akutsu81}, pure POPC data from \cite{altenbach84}, 
%    POPC:POPS mixture data from \cite{roux90}, POPC:POPG mixture data with 0.1M NaCl from \cite{macdonald87}
%    and POPC:POPG mixture data without NaCl from \cite{borle85}.
%  }
%  \todo{Check the NaCl concentrations in the samples.}
%\end{figure}
%\begin{figure}[]
%  \centering
%  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaClBELOW1M.eps}
%  \caption{\label{OrderParametersWithCaClBELOW1M}
%    Figure \ref{OrderParametersWithCaCl} zoomed to smaller concentrations.
%  }
%\end{figure}
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaCl.eps}
  \caption{\label{OrderParameterCHANGESWithCaClBELOW1M}
    The change of PC headgroup order parameters
    in the presence of different amount of negatively charged lipids
    respect to the values without added CaCl$_2$.
    The original data is the same as in Fig. \ref{OrderParametersWithCaCl}. 
    Pure DPPC data from \cite{akutsu81}, pure POPC data from \cite{altenbach84}, 
    POPC:POPS mixture data from \cite{roux90}, POPC:POPG mixture data with 0.1M NaCl from \cite{macdonald87}
    and POPC:POPG mixture data without NaCl from \cite{borle85}.
  }
\end{figure}

>>>>

Based on the subjective ranking shown in Fig. \ref{comparisonTablePS},
the CHARMM36 is the best performing model for both lipids.
However, none of the tested models give a satisfactory 
agreement with experiments for the headgroup order paramaters of 
PS lipids. The total deviation from the experiments ($\Sigma$ in Fig. \ref{comparisonTablePS})
for the best performing CHARMM36 model is larger for PS lipids (8)
than for PC (3) \cite{botan15}. Therefore the 
interpretation of structural details of PS headgroup or
differences between PC and PS lipids is very challenging 
from the current MD simulation models.
Figures \ref{dihedralsGLY} and \ref{dihedralsHG} show dihedral
angle distributions calculated from different models. 
The glycerol backbone structures from CHARMM36 and
Slipids simulations visualized in Fig. \ref{glycerol_buslaev} reveal the
missing structures in Slipids model, which probably lead to the incorrect
order parameters. For the headgroup $\alpha$ and $\beta$ carbon order parameters of PS lipids,
the tested models perform less well than for PC headgroup in previous
study \cite{botan15}.
\todo{Discussion will be finished when we have all the data in
  Figs. \ref{HGorderParametersPS}, \ref{dihedralsGLY} and \ref{dihedralsHG}.
  One possible conclusion could be the following:
  The main differences between the models in the headgroup region are observed
  for dihedrals C12-C11-O12-P and C11-C12-C13-O1A.
  CHARMM36, CHARMM36UA and Slipids give very similar results to the
  dihedral C11-C12-C13-O1A, which is close to the $\beta$-carbon.
  The order parameters of $\beta$-carbons for these three models
  are in best argeement with the experiments in figure \ref{HGorderParametersPS}.
  On the other hand, Gromos-CKP models give better order parameters
  for $\alpha$-carbon than Slipids, CHARMM36 or CHARMM36UA. 
  In conclusion, the suggestion would be that the single peak for
  observed at 120 degrees in CHARMMs and Slipids would
  be more realistic for C11-C12-C13-O1A dihedral, while the single peak at 180 degrees observed
  in CKP models and in Berger would be most realistic for C12-C11-O12-P dihedral.
}

@@

The headgroup order parameters of PC lipids
decrease proportionally to the bound positive
charge in to a bilayer \cite{seelig87,catte16} and can be
therefore used to measure the ion binding
affinity. This molecular electrometer concept can
be also applied to lipid bilayers with mixtures
of PC and negatively charged lipids \cite{borle85,macdonald87,roux90}
(see Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}).
%This is demonstrated in Fig. %s \ref{OrderParametersWithCaCl},
%\ref{OrderParametersWithCaClBELOW1M} and
%using the previuosly reported experimental data
%for mixtures with varying proportions of
%negatively charged PS or PG lipids.
%
%the changes of PC headgroup order parameters %for PC headgroup $\alpha$ and $\beta$ carbons
%as a function of CaCl$_2$ concentration are shown
%
% (see Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}).

@@

%\begin{figure}[]
%  \centering
%  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaCl.eps}
%  \caption{\label{OrderParametersWithCaCl}
%    PC headgroup order parameters as a function of CaCl concentration from experiments containing charged lipids.
%    Pure DPPC data from \cite{akutsu81}, pure POPC data from \cite{altenbach84}, 
%    POPC:POPS mixture data from \cite{roux90}, POPC:POPG mixture data with 0.1M NaCl from \cite{macdonald87}
%    and POPC:POPG mixture data without NaCl from \cite{borle85}.
%  }
%  \todo{Check the NaCl concentrations in the samples.}
%\end{figure}
%\begin{figure}[]
%  \centering
%  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaClBELOW1M.eps}
%  \caption{\label{OrderParametersWithCaClBELOW1M}
%    Figure \ref{OrderParametersWithCaCl} zoomed to smaller concentrations.
%  }
%\end{figure}
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaCl.eps}
  \caption{\label{OrderParameterCHANGESWithCaClBELOW1M}
    The change of PC headgroup order parameters as a function of CaCl$_2$
    measured from bilayers containing different amount of negatively charged lipids.
    The values are taken from 2H NMR experiments reported in the
    literature (DPPC \cite{akutsu81}, POPC \cite{altenbach84}, POPC:POPS (5:1) \cite{roux90},
    POPC:POPG  mixtures with 0.1M NaCl \cite{macdonald87}
    and POPC:POPG (1:1) without NaCl \cite{borle85}).
    As expected, the decrease of order parameters with the added CaCl$_2$
    is more pronounced for systems with larger fraction of
    negatively charged lipids, indicating larger amount amount
    of bound cations.
  }
\end{figure}
